---
title: "flight_model"
author: "Group 5"
output: html_document
date: "2025-03-23"
---

```{r}
# load necessary libraries
library(tidyverse)
library(brms)  # bayesian regression
library(ggplot2)
library(scales)
library(caret)
library(tidyverse)
```

```{r}
# read data
flight_data <- read.csv("flight_data_cleaned.csv")
# delete rows containing NA in the x variables
flight_data <- flight_data %>%
  drop_na(Fare, Number.Of.Stops, Total_Minutes, distance,
          IsWeekend, ifHoliday, Is_Low_Cost, Low_Cost_Count,
          Departure.Off.Peak, Arrival.Off.Peak)

mean_fare = mean(flight_data$Fare)
sd_fare = sd(flight_data$Fare)
cat(mean_fare,"\n")
cat(sd_fare,"\n")
```


# 1. Basic models

## 1.1 linear regression

```{r}
# basic linear regression model
lm_basic <- lm(Fare ~ Number.Of.Stops + Total_Minutes + distance + 
               IsWeekend + ifHoliday + Is_Low_Cost + Low_Cost_Count + 
               Departure.Off.Peak + Arrival.Off.Peak,
               data = flight_data)

# summary
summary(lm_basic)
```

## 1.2  bayesian regression

```{r}
# basic bayesian regression model
bayes_model <- brm(
  Fare ~ Number.Of.Stops + Total_Minutes + distance + 
  IsWeekend + ifHoliday + Is_Low_Cost + Low_Cost_Count + 
  Departure.Off.Peak + Arrival.Off.Peak,
  data = flight_data,
  family = gaussian(),
  chains = 4,
  iter = 2000
)
```

```{r}
# summary of bayesian regression model and bayesian R squared
summary(bayes_model)
bayes_R2(bayes_model)
```


# 2. Models with interaction terms

## 2.1 weekend and holiday

```{r}
# linear regression model with interaction term 1
lm_interaction1 <- lm(Fare ~ Number.Of.Stops + Total_Minutes + distance + 
                    IsWeekend + ifHoliday + Is_Low_Cost + Low_Cost_Count + 
                    Departure.Off.Peak + Arrival.Off.Peak+
                    IsWeekend:ifHoliday, # weekend and holiday
                    data = flight_data)

# compare model with interaction1 with basic model
anova(lm_basic, lm_interaction1)
```

## 2.2 low-cost operators and distance

```{r}
# linear regression model with interaction term 2
lm_interaction2 <- lm(Fare ~ Number.Of.Stops + Total_Minutes + distance + 
                    IsWeekend + ifHoliday + Is_Low_Cost + Low_Cost_Count + 
                    Departure.Off.Peak + Arrival.Off.Peak+
                    Is_Low_Cost:distance,  # Differences in pricing strategies of low-cost operators across various distances
                    data = flight_data)

# compare model with interaction2 with basic model
anova(lm_basic, lm_interaction2)

# Visualize the pricing differences of low-cost operators across various distances
ggplot(flight_data, aes(x = distance, y = Fare, color = factor(Is_Low_Cost))) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm") +
  theme_minimal() +
  labs(title = "The distance-Fare relationship of Low-cost operators vs. regular operators",
       color = "whether low-cost operator or not(0=no, 1=yes)")
```

## 2.3 Number.Of.Stops and Total_Minutes

```{r}
# linear regression model with interaction term 3
lm_interaction3 <- lm(Fare ~ Number.Of.Stops + Total_Minutes + distance + 
                    IsWeekend + ifHoliday + Is_Low_Cost + Low_Cost_Count + 
                    Departure.Off.Peak + Arrival.Off.Peak+
                    Total_Minutes:Number.Of.Stops,  # an increase in the Number.Of.Stops on Total_Minutes affect Fare
                    data = flight_data)

# compare model with interaction3 with basic model
anova(lm_basic, lm_interaction3)

# Analyze the impact of Total_Minutes on Fare under different Number.Of.Stops
ggplot(flight_data, aes(x = Total_Minutes, y = Fare, color = factor(Number.Of.Stops))) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm") +
  theme_minimal() +
  labs(title = "Total_Minutes-Fare relationship under different Number.Of.Stops",
       color = "Number.Of.Stops")
```

## 2.4 Departure.Off.Peak and Arrival.Off.Peak

```{r}
# linear regression model with interaction term 4
lm_interaction4 <- lm(Fare ~ Number.Of.Stops + Total_Minutes + distance + 
                    IsWeekend + ifHoliday + Is_Low_Cost + Low_Cost_Count + 
                    Departure.Off.Peak + Arrival.Off.Peak+
                    Departure.Off.Peak:Arrival.Off.Peak,  # The combined effect of whether departure and arrival times are during peak hours
                    data = flight_data)

# compare model with interaction4 with basic model
anova(lm_basic, lm_interaction4)

```

## 2.5 Total_Minutes and ifHoliday

```{r}
# linear regression model with interaction term 5
lm_interaction5 <- lm(Fare ~ Number.Of.Stops + Total_Minutes + distance + 
                    IsWeekend + ifHoliday + Is_Low_Cost + Low_Cost_Count + 
                    Departure.Off.Peak + Arrival.Off.Peak+
                    Total_Minutes:ifHoliday,  # Whether IfHoliday plays as moderator between totalMinutes and Fare
                    data = flight_data)

# compare model with interaction5 with basic model
anova(lm_basic, lm_interaction5)
```

## 2.6 Total_Minutes and IsWeekend

```{r}
# linear regression model with interaction term 6
lm_interaction6 <- lm(Fare ~ Number.Of.Stops + Total_Minutes + distance + 
                    IsWeekend + ifHoliday + Is_Low_Cost + Low_Cost_Count + 
                    Departure.Off.Peak + Arrival.Off.Peak+
                    Total_Minutes:IsWeekend, # Whether IsWeekend plays as moderator totalMinutes and Fare
                    data = flight_data)

# compare model with interaction6 with basic model
anova(lm_basic, lm_interaction6)
```

## 2.7 Departure.Off.Peak and Number.Of.Stops

```{r}
# linear regression model with interaction term 7
lm_interaction7 <- lm(Fare ~ Number.Of.Stops + Total_Minutes + distance + 
                    IsWeekend + ifHoliday + Is_Low_Cost + Low_Cost_Count + 
                    Departure.Off.Peak + Arrival.Off.Peak+
                    Departure.Off.Peak:Number.Of.Stops,  # passengers may tolerate more stops duting day flights compared to night flights 
                    data = flight_data)

# compare model with interaction7 with basic model
anova(lm_basic, lm_interaction7)
```

## 2.8 comprehensive model with effective interaction terms

### 2.8.1 linear regression

```{r}
# construct a comprehensive linear regression model including important interaction terms
lm_important_interactions <- lm(Fare ~ Number.Of.Stops + Total_Minutes + distance + 
                    IsWeekend + ifHoliday + Is_Low_Cost + Low_Cost_Count + 
                    Departure.Off.Peak + Arrival.Off.Peak +
                    Is_Low_Cost:distance +  # interaction2
                    Total_Minutes:Number.Of.Stops +  # interaction3
                    Departure.Off.Peak:Arrival.Off.Peak,  # interaction4
                    data = flight_data)

# summary
summary(lm_important_interactions)
```

### 2.8.2 bayesian regression

```{r}
# construct a comprehensive bayesian regression model including important interaction terms
bayes_important_interactions <- brm(
  Fare ~ Number.Of.Stops + Total_Minutes + distance + 
  IsWeekend + ifHoliday + Is_Low_Cost + Low_Cost_Count + 
  Departure.Off.Peak + Arrival.Off.Peak +
  Is_Low_Cost:distance +
  Total_Minutes:Number.Of.Stops +
  Departure.Off.Peak:Arrival.Off.Peak,
  data = flight_data,
  family = gaussian(),
  chains = 4,
  iter = 2000
)

# summary and bayesian R squared
summary(bayes_important_interactions)
bayes_R2(bayes_important_interactions)
```


# 3. Cross validation & model improvement

## 3.1 10-fold cv with un-scaled data

```{r}
# reproducibility
set.seed(123)

# define the model formula
model_formula <- Fare ~ Number.Of.Stops + Total_Minutes + distance + 
                IsWeekend + ifHoliday + Is_Low_Cost + Low_Cost_Count + 
                Departure.Off.Peak + Arrival.Off.Peak +
                Is_Low_Cost:distance +
                Total_Minutes:Number.Of.Stops +
                Departure.Off.Peak:Arrival.Off.Peak

# define 10 fold cv
train_control <- trainControl(
  method = "cv",
  number = 10,  # 10-fold CV
  verboseIter = TRUE
)
# cv
cv_model <- train(
  model_formula,
  data = flight_data,
  method = "lm",
  trControl = train_control
)

# print the result
print(cv_model)

# get more detailed prediction metrics
print(cv_model$results)

# visualize the prediction results
# use the final model to predict 
predictions <- predict(cv_model, flight_data)

# scatter plot of prediction values vs. actual values
ggplot(data.frame(predicted = predictions, actual = flight_data$Fare), 
       aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "prediction values vs actual values",
    x = "actual fare",
    y = "prediction fare"
  )
cv_metrics <- cv_model$resample
print(cv_metrics)

# calculate Mean_RMSE/SD_RMSE/Mean_Rsquared/SD_RSquared
summary_metrics <- data.frame(
  Mean_RMSE = mean(cv_metrics$RMSE),
  SD_RMSE = sd(cv_metrics$RMSE),
  Mean_Rsquared = mean(cv_metrics$Rsquared),
  SD_Rsquared = sd(cv_metrics$Rsquared)
)
print(summary_metrics)
```

## 3.2 10-fold cv with scaled data

Why scale?

To make our RMSE range between 0-1 to evaluate the model more straightforward

```{r}
# check the satistic values of "Fare"
mean_fare = mean(flight_data$Fare)
sd_fare = sd(flight_data$Fare)
cat("Fare's mean：", mean_fare, "\n")
cat("Fare's standard deviation：", sd_fare, "\n")

# create data after scaling
flight_data_scaled <- flight_data %>%
  mutate(Fare_scaled = scale(Fare))

# use scaled data for cv
set.seed(123)

train_control <- trainControl(
  method = "cv",
  number = 10,
  verboseIter = TRUE
)

# use scaled data to create model
cv_model_scaled <- train(
  Fare_scaled ~ Number.Of.Stops + Total_Minutes + distance + 
        IsWeekend + ifHoliday + Is_Low_Cost + Low_Cost_Count + 
        Departure.Off.Peak + Arrival.Off.Peak +
        Is_Low_Cost:distance +
        Total_Minutes:Number.Of.Stops +
        Departure.Off.Peak:Arrival.Off.Peak,
  data = flight_data_scaled,
  method = "lm",
  trControl = train_control
)

print(cv_model_scaled)

# visualize prediction results
predictions_scaled <- predict(cv_model_scaled, flight_data_scaled)

ggplot(data.frame(predicted = predictions_scaled, 
                  actual = flight_data_scaled$Fare_scaled), 
       aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "predictive fare VS actual fare after scaling",
    x = "actual fare - scaled",
    y = "predictive fare - scaled"
  )
```

## 3.3 model improvement

Why improve?

The RMSE is kind of big, meaning that our model cannot explain the variation very well.

```{r}
# set seed and cv parameters
set.seed(123)
train_control <- trainControl(
  method = "cv",
  number = 10,
  verboseIter = TRUE
)
```

### 3.3.1 add non linear terms

```{r}
# solution 1: add non linear terms
model_nonlinear <- train(
  Fare_scaled ~ Number.Of.Stops + Total_Minutes + I(Total_Minutes^2) + 
        distance + I(distance^2) +  # add squared terms
        IsWeekend + ifHoliday + Is_Low_Cost + Low_Cost_Count + 
        Departure.Off.Peak + Arrival.Off.Peak +
        Is_Low_Cost:distance +
        Total_Minutes:Number.Of.Stops +
        Departure.Off.Peak:Arrival.Off.Peak,
  data = flight_data_scaled,
  method = "lm",
  trControl = train_control
)

print("Results of non-linear model: ")
print(model_nonlinear$results)
```
### 3.3.2 Random Forest

```{r}
# Solution 2: Random Forest
library(randomForest)
model_rf <- train(
  Fare_scaled ~ Number.Of.Stops + Total_Minutes + distance + 
        IsWeekend + ifHoliday + Is_Low_Cost + Low_Cost_Count + 
        Departure.Off.Peak + Arrival.Off.Peak,
  data = flight_data_scaled,
  method = "rf",
  trControl = train_control,
  ntree = 500,
  importance = TRUE
)

print("Results of random forest: ")
print(model_rf$results)

# check feature importance
importance_rf <- varImp(model_rf)
print("feature importance")
print(importance_rf)
```

### 3.3.3 Feature Engineering

```{r}
# Solution 3: Feature Engineering
flight_data_engineered <- flight_data_scaled %>%
  mutate(
    # create new features
    peak_hours = ifelse(Departure.Off.Peak == 0 | Arrival.Off.Peak == 0, 1, 0),
    distance_per_minute = distance/Total_Minutes,
    total_cost_factor = Is_Low_Cost * Low_Cost_Count,
    weekend_holiday = IsWeekend * ifHoliday,
    stops_per_distance = Number.Of.Stops/distance
  )

model_engineered <- train(
  Fare_scaled ~ Number.Of.Stops + Total_Minutes + distance + 
        IsWeekend + ifHoliday + Is_Low_Cost + Low_Cost_Count + 
        Departure.Off.Peak + Arrival.Off.Peak +
        peak_hours + distance_per_minute + total_cost_factor + 
        weekend_holiday + stops_per_distance,
  data = flight_data_engineered,
  method = "lm",
  trControl = train_control
)

print("Results of feature engineering: ")
print(model_engineered$results)
```

### 3.3.4 Hierarchical Bayesian Model

```{r}
# Solution 4：Hierarchical Bayesian Model

# create hierarchical Bayesian Model
bayes_model <- brm(
  Fare_scaled ~ Number.Of.Stops + Total_Minutes + distance + 
  IsWeekend + ifHoliday + Is_Low_Cost + Low_Cost_Count + 
  Departure.Off.Peak + Arrival.Off.Peak +
  Is_Low_Cost:distance +
  Total_Minutes:Number.Of.Stops +
  Departure.Off.Peak:Arrival.Off.Peak +
  # Modify the hierarchical structure to retain only one random effect
  (distance | Is_Low_Cost),  # Allow different distance effects for different airline types
  data = flight_data_scaled,
  family = gaussian(),
  chains = 4,
  iter = 2000,
  cores = 4
)

summary(bayes_model)

predictions_bayes <- predict(bayes_model, flight_data_scaled)[,"Estimate"]

# Calculate Bayes R2
bayes_r2 <- bayes_R2(bayes_model)
print("bayes R²: ")
print(bayes_r2)
```




### 3.3.5 Compare three solutions

```{r}
# compare the performance of three models
results_comparison <- data.frame(
  Model = c("non-linear model", "random forest", "feature engineering","hierarchical bayesian model"),
  RMSE = c(
    min(model_nonlinear$results$RMSE),
    min(model_rf$results$RMSE),
    min(model_engineered$results$RMSE),
    sqrt(mean((predictions_bayes - flight_data_scaled$Fare_scaled)^2))
  ),
  Rsquared = c(
    max(model_nonlinear$results$Rsquared),
    max(model_rf$results$Rsquared),
    max(model_engineered$results$Rsquared),
    mean(bayes_r2[,1])
  )
)

print("model comparison: ")
print(results_comparison)
```


```{r}
# compare by visualization

# create prediction values for each model
predictions_nonlinear <- predict(model_nonlinear, flight_data_scaled)
predictions_rf <- predict(model_rf, flight_data_scaled)
predictions_engineered <- predict(model_engineered, flight_data_engineered)
predictions_bayes <- predict(bayes_model, flight_data_scaled)[,"Estimate"]

# create comparison plots of prediction values
library(ggplot2)
library(gridExtra)

p1 <- ggplot(data.frame(actual = flight_data_scaled$Fare_scaled, 
                        predicted = predictions_nonlinear), 
             aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  theme_minimal() +
  labs(title = "performance of non-linear model")

p2 <- ggplot(data.frame(actual = flight_data_scaled$Fare_scaled, 
                        predicted = predictions_rf), 
             aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  theme_minimal() +
  labs(title = "performance of random forest")

p3 <- ggplot(data.frame(actual = flight_data_scaled$Fare_scaled, 
                        predicted = predictions_engineered), 
             aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  theme_minimal() +
  labs(title = "performance of feature engineering")

p4 <- ggplot(data.frame(actual = flight_data_scaled$Fare_scaled, 
                        predicted = predictions_bayes), 
             aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  theme_minimal() +
  labs(title = "Hierarchical Bayesian model prediction performance")

grid.arrange(p1, p2, p3,p4, ncol = 2)
```


